Containers and R: the Rockerverse and beyond

### 2019-09-09 manuscript.Rmd

---
title: "Containers and R: the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: Institute for Geoinformatics, University of Münster
    address:
    - Heisenbergstr. 2, 48149 Münster, Germany
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    #affiliatin: -
    #email: -
  - name: Dirk Eddelbuettel
    #affiliation:
    #email: -
  - name: Mark Edmondson
    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
    email: mark@markedmondson.me
  - name: Colin Fay
    affiliation: ThinkR
    email: contact@colinfay.me
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Karthik Ram
    #affiliation: -
    #email: -
  - name: Noam Ross
    #affiliation: -
    #email: -
  - name: Nan Xiao
    #affiliation: -
    #email: -
  - name: Lori Shepherd
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
    email: lori.shepherd@roswellpark.org
  - name: Nitesh Turaga
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
    email: nitesh.turaga@roswellpark.org
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
preamble: >
  \usepackage{longtable}
  \usepackage{hyperref}
output: rticles::rjournal_article
#  includes:
#    in_header: preamble.tex
---

@@

- [`containerit`](https://github.com/o2r-project/containerit/)
- [`dockertest`](https://github.com/traitecoevo/dockertest/)
- [`liftr`](https://liftr.me/) for OS-level reproducibility [@nanx]
- [`rize` for Shiny](https://github.com/cole-brokamp/rize)
- [`dockerfiler`](https://github.com/ColinFay/dockerfiler)

>>>>

---
title: "Containers and R: the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: Institute for Geoinformatics, University of Münster
    address:
    - Heisenbergstr. 2, 48149 Münster, Germany
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    #affiliatin: -
    #email: -
  - name: Dirk Eddelbuettel
    #affiliation:
    #email: -
  - name: Mark Edmondson
    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
    email: mark@markedmondson.me
  - name: Colin Fay
    affiliation: ThinkR
    email: contact@colinfay.me
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Karthik Ram
    #affiliation: -
    #email: -
  - name: Noam Ross
    #affiliation: -
    #email: -
  - name: Nan Xiao
    affiliation: Seven Bridges Genomics
    address: "529 Main St, Suite 6610 \\newline Charlestown, MA 02129, USA \\newline ORCiD: 0000-0002-0250-5673"
    email: me@nanx.me
  - name: Lori Shepherd
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
    email: lori.shepherd@roswellpark.org
  - name: Nitesh Turaga
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
    email: nitesh.turaga@roswellpark.org
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
preamble: >
  \usepackage{longtable}
  \usepackage{hyperref}
output: rticles::rjournal_article
#  includes:
#    in_header: preamble.tex
---

@@

- [`containerit`](https://github.com/o2r-project/containerit/)
- [`dockertest`](https://github.com/traitecoevo/dockertest/)

### liftr

[_liftr_](https://nanx.me/liftr/) \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
Currently, the R Markdown format and its backend compilation engine _knitr_ offer a _de facto_ standard for creating dynamic documents \citep{xie2018}.
However, the reproducibility of such content authoring environments is often limited to individual machines --- it is not easy to replicate the system environment (libraries, R versions, R packages) where the document was compiled.
This issue becomes even more serious when it comes to collaborative document authoring and creating large-scale document building services.
_liftr_ solves this reproducibility problem by bringing Docker to the game.
In essence, _liftr_ helps R Markdown users create and manage Docker containers for rendering the documents, thus make the computations utterly reproducible across machines and systems.

On implementation, with no side effects, _liftr_ extended and introduced new metadata fields to R Markdown, allowing users to declare the dependencies for rendering the document.
_liftr_ parses such fields and generates a `Dockerfile` for creating Docker containers.
_liftr_ then helps render the document inside the created Docker container.
This workflow is summarized in Figure \ref{figure:liftr}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{liftr-workflow}
  \caption{The liftr workflow for rendering containerized R Markdown documents.}
  \label{figure:liftr}
\end{figure}

_knitr_ and R Markdown are used as the template engine to generate the `Dockerfile`.
Features such as caching container layers for saving image build time, automatic housekeeping for fault-tolerant builds, and Docker status check are supported by _liftr_.
Four RStudio addins are also offered by _liftr_ to allow push-button compilation of documents and provide better IDE integrations.

Three basic principles are followed to design the _liftr_ package since its inception.

1. Continuous reproducibility.
Continuous integration, continuous delivery, and continuous deployment are well-accepted practices in software engineering.
Similarly, it is believed by the authors of _liftr_ that ensuring computational reproducibility means a continuous process instead of creating static data/code archives or a one-time deal.
Specifically, the software packages used in data analysis should be upgraded regularly in a manageable way.
Therefore, _liftr_ supports specifying particular versions of package dependencies, while users are encouraged to always use the latest version of packages (without a version number) by default.

2. Document first.
Many data analysis workflows could be wrapped as either R packages or dynamic documents.
In _liftr_, the endpoint of dynamic report creation is the focus of containerization, because this offers more possibilities for organizing both computations and documentation.
Users are encouraged to start thinking from the visible research output from the first day.

3. Minimal footprint.
R Markdown and Docker are already complex software systems.
Making them work together seamlessly can be complicated.
Therefore, API designs such as function arguments are simplified while being kept as expressive and flexible as possible.

In summary, _liftr_ tries to redefine the meaning of computational reproducibility by offering system-level reproducibility for data analysis.
It provided a practical way for achieving it --- a new perspective on how reproducible research could be done in reality.
Further, sharing system environments for data analysis also becomes extremely easy, since users only need to share the R Markdown document (with a few extra metadata fields), and compile them with _liftr_.
As an example, _liftr_ demonstrated its advantage for R Markdown-based computational workflow orchestration, by effortlessly containerizing 18 complex Bioconductor workflows in the DockFlow project (https://dockflow.org) in 2017.

- [`rize` for Shiny](https://github.com/cole-brokamp/rize)
- [`dockerfiler`](https://github.com/ColinFay/dockerfiler)
